---
title: "Data formats in package breathtestcore"
author: "Dieter Menne,  dieter.menne@menne.biomed.de"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Concepts

<sup>13</sup>C data can be imported in generic formats stored in Excel files, and in several vendor-specific formats, e.g. from BreathID and Wagner/IRIS. A collection of sample files with and without errors is available in the directory ``r R.home()`/library/breathtestcore/extdata`; function `btcore_file()` retrieves the names and long path of the available data sets. 

```{r, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(comment = NA)
```

```{r}
library(breathtestcore)
head(btcore_file())
btcore_file("Standard.TXT")
```


When you know the format, you can read in the data using the special functions, e.g. `read_breathid()`. 

When you do not know the format, or when you want to read several different file formats at once, use function `read_any_breathtest()` which tries to guess the format. 

```{r}
files = c(
  btcore_file("IrisCSV.TXT"), # Wagner/IRIS format
  btcore_file("350_20043_0_GER.txt") # BreathID
 )
 bt = read_any_breathtest(files)
 # Returns a list of elements of class breathtest_data
 str(bt, 1)
 bt_df = cleanup_data(bt)
 str(bt_df)
```
Passing through `cleanup_data()` returns a data frame/tibble and adds a grouping variable. 

To plot data without fitting, use `null_fit()`.


```{r, fig.height = 2, fig.width =4}
 nf = null_fit(bt_df)
 str(nf)
 
 plot(nf) # dispatches to plot.breathtestfit
```

To add new formats, override  `breathtest_read_function()` and add a new function that returns a structure given by `breathtest_data()`.

> Always pass data through function `cleanup_data()` to obtain a data frame that can be fed to one of the fitting functions `nls_fit()`, `nlme_fit()`, `null_fit()` or `breathteststan::stan_fit()`.

## Automatic grouping

You can add a grouping variable, e.g. for multiple meal types, to compute between group differences of means. Cross-over, randomized or mixed designs (some patients cross-over) are supported.

You must explicitlty state the grouping variable for each single file as shown below. Without names, it is possible to vectorize, e.g. `read_any_breathtest(c(file1, file2))`, but the 'c()' operator used on vectors disambiguates the names by appending numbers.   

```{r, fig.height = 2.5, fig.width = 8}
files1 = c(
  group_a = btcore_file("IrisCSV.TXT"), # Use only single file with grouping
  group_a = btcore_file("Standard.TXT"),
  group_b = btcore_file("350_20043_0_GER.txt")
 )

# Alternative syntax using magrittr operator
suppressPackageStartupMessages(library(dplyr))
read_any_breathtest(files1) %>% 
  cleanup_data() %>% 
  null_fit() %>% 
  plot()
```


## Simulated data

Function `simulate_breathtest_data` generates sample data to test different algorithms. Curves with outliers can be generated by setting `student_t_df` to values from 2 (very strong outliers) to 10 (almost gaussian). 

```{r,  fig.height = 5, fig.width = 6, fig.cap = "Example of a cross-over design with missing data, outliers and missing record in the red curve."}
set.seed(212)
data = list(meal_a = simulate_breathtest_data(n_records = 3, noise = 2,
                          student_t_df = 3, missing = 0.3), 
            meal_b = simulate_breathtest_data(n_records = 4))
data %>% 
  cleanup_data() %>% 
  nlme_fit() %>% 
  plot()

```

```{r, fig.cap= "Function simulate_breathtest_data returns the values of the parameters used to generate the data. These can be used to check the results of the model prediction."}
data$meal_a$record
```

## Built-in data sets

Three data sets are included in R format and can be loaded as shown below. All data were provided by the University Hospital of Zürich; details are given in the documentation.

```{r}
data("usz_13c")
cat("usz_13c has data from", length(unique(usz_13c$patient_id)), "patients with" , 
    length(unique(usz_13c$group)), "different meals")
```

* `breathtestcore::usz_13c` A large data set used to establish reference ranges for healthy volunteers and patients
* `breathtestcore::usz_13c_a` Exotic data, a challenge for fitting algorthms
* `breathtestcore::usz_13c_d` Has gastric emptying half time from MRI as attribute, and can used to compare recorded data with gold standards; see the example in the documentation.


